---
title: "DATA606_os4_readings"
author: "Bonnie Cooper"
date: "2/8/2020"
output:
  html_document:
    css: ./lab.css
    highlight: pygments
    theme: cerulean
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
<br><br>

## OpenIntro Statistics Reading Notes
<br><br>

```{r, echo = FALSE}
library( tidyverse )
library( dplyr )
library( ggplot2 )
```


### Chapter 2: Summarizing Data
<br>
This Chapter focuses on the mechanics of generating summary statistics and summary graphs of data in R
<br>

#### 2.1 Examining Numerical Data
<br>

summarizing numerical variables using the loan50 and county datasets

```{r}
#read the .csv files from URLs hosted on github into R dataframes 
countyURL <- 'https://raw.githubusercontent.com/SmilodonCub/DATA606/master/county.csv'
loan50URL <- 'https://raw.githubusercontent.com/SmilodonCub/DATA606/master/loan50.csv'
county <- read.csv( url( countyURL ) )
loan50 <- read.csv( url( loan50URL ) )
```

##### Scatterplots for paired data

case-by-case view of data for numerical variables

```{r}
data2plot <- loan50 %>% select( Income = total_income, Loan = loan_amount )
ggplot(data2plot, aes(x=Income, y=Loan)) +
    geom_point(shape=1) +    # Use hollow circles
    geom_smooth()            # Add a loess smoothed fit curve with confidence region
#> `geom_smooth()` using method = 'loess'
```

The relationship between variables in these two graphs is NONLINEAR: shows relationships best described with curvature rather than a straight line.

```{r}
data2plot <- county %>% select( Pov_Rate = poverty, Income = median_hh_income )
ggplot(data2plot, aes(x=Pov_Rate, y=Income)) +
    geom_point(shape=1) +    # Use hollow circles
    geom_smooth()            # Add a loess smoothed fit curve with confidence region
#> `geom_smooth()` using method = 'loess'
```

#### Dot plots for single variables.
Mean is a common ay to measure the center of a distribution
Sum(values)/num(values)
the sample mean can be computed as the sum of the observed values divided by the number of observations.
mu represents the average of all observations in the population; the sample mean provides a rough estimate of mu. the sample mean is a point estimate; the larger the sample, the better the estimate.


