---
title: "Chapter 8 - Introduction to Linear Regression"
author: ""
output:
    pdf_document:
        extra_dependencies: ["geometry", "multicol", "multirow", "xcolor"]
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

**Nutrition at Starbucks, Part I.** (8.22, p. 326) The scatterplot below shows the relationship between the number of calories and amount of carbohydrates (in grams) Starbucks food menu items contain. Since Starbucks only lists the number of calories on the display items, we are interested in predicting the amount of carbs a menu item has based on its calorie content.

```{r, echo=FALSE, message=FALSE, warning=FALSE, fig.show="hold", out.width="33%", fig.height=4}
library(openintro)
# load data ---------------------------------------------------------
starbucks <- read.csv("https://github.com/jbryer/DATA606Fall2019/raw/master/course_data/starbucks.csv")
# model calories vs. carbos -----------------------------------------
m_carb_cals <- lm(carb ~ calories, data = starbucks)
# plot calories vs. carbos ------------------------------------------
par(mar = c(3.5, 4, 1, 0.5), las = 1, mgp = c(2.5, 0.7, 0), 
    cex.lab = 1.5, cex.axis = 1.5)
plot(carb ~ calories, data = starbucks, 
     pch = 19, col = COL[1,2], 
     xlab = "Calories", ylab = "Carbs (grams)", axes = FALSE)
axis(1)
axis(2, at = seq(20, 80, 20))
box()
abline(m_carb_cals, col = COL[2], lwd = 2)
# plot residuals ----------------------------------------------------
par(mar = c(3.5, 4, 1, 0.5), las = 1, mgp = c(2.5, 0.7, 0), 
    cex.lab = 1.5, cex.axis = 1.5)

plot(m_carb_cals$residuals ~ starbucks$calories,
     xlab = "Calories", ylab = "Residuals", 
     col = COL[1,2], pch = 19,
     ylim = c(-30, 30), axes = FALSE)
axis(1)
axis(2, at = seq(-20, 20, 20))
box()
abline(h = 0, lty = 2)
# histogram of residuals --------------------------------------------
par(mar = c(3.5, 2.5, 0.5, 0.5), las = 1, mgp = c(2.5, 0.7, 0), 
    cex.lab = 1.5, cex.axis = 1.5)

hist(m_carb_cals$residuals,
     col = COL[1], 
     xlab = "Residuals", ylab = "", main = "", 
     axes = FALSE, xlim = c(-40,40))
axis(1, at = seq(-40, 40, 20))
axis(2)
```

(a) Describe the relationship between number of calories and amount of carbohydrates (in grams) that Starbucks food menu items contain.  

Calories and the amount of carbohydrates for Starbucks menu items appear to have a resonably linear positive relationship, i.e. as the number of calories increase, so does the amount of carbs.

(b) In this scenario, what are the explanatory and response variables?  

In this scenario, the explanatory variable is Calories and the amount of Carbs(g) is the response variable.  

(c) Why might we want to fit a regression line to these data?  

W might want to fit a regression line so that we could interpolate Carb estimates for a product given we know the number of Calories.  


(d) Do these data meet the conditions required for fitting a least squares line?  

No, these data do notmeet the criteria for a least squares line, because the condition of Constant variability is not met, because there is a relationship between the variability of carbs and calories (the variability of carbs is larger when calories is higher).





--------------------------------------------------------------------------------

\clearpage

**Body measurements, Part I.** (8.13, p. 316) Researchers studying anthropometry collected body girth measurements and skeletal diameter measurements, as well as age, weight, height and gender for 507 physically active individuals. The scatterplot below shows the relationship between height and shoulder girth (over deltoid muscles), both measured in centimeters.

\begin{center}
```{r, echo=FALSE, message=FALSE, warning=FALSE, fig.show="hold", out.width="50%", fig.height=4}
# load packages -----------------------------------------------------
library(openintro)
# load data ---------------------------------------------------------
data(bdims)
# plot height vs. shoulder girth ------------------------------------
par(mar = c(3.8, 3.8, 0.5, 0.5), las = 1, mgp = c(2.7, 0.7, 0), 
    cex.lab = 1.25, cex.axis = 1.25)
plot(bdims$hgt ~ bdims$sho.gi, 
     xlab = "Shoulder girth (cm)", ylab = "Height (cm)", 
     pch = 19, col = COL[1,2])
```
\end{center}

(a) Describe the relationship between shoulder girth and height.  

Shoulder girth and height appear to have a resonably linear positive relationship, i.e. as the height increases, so does shoulder girth.

(b) How would the relationship change if shoulder girth was measured in inches while the units of height remained in centimeters?  

The general relationship would not change, this would only alter the units the data is plotted in and could be easily converted and replotted to express the same relationship in the given figure. 




--------------------------------------------------------------------------------

\clearpage
                                      
**Body measurements, Part III.** (8.24, p. 326) Exercise above introduces data on shoulder girth and height of a group of individuals. The mean shoulder girth is 107.20 cm with a standard deviation of 10.37 cm. The mean height is 171.14 cm with a standard deviation of 9.41 cm. The correlation between height and shoulder girth is 0.67.

(a) Write the equation of the regression line for predicting height.  

```{r}
#find the slope
#slope = sd_height/sd_girth*R
sdh <- 9.41
sdg <- 10.37
R <- 0.67
b_1 <- sdh/sdg*R
#use the feature means to find the intercept
mh <- 171.14
mg <- 107.20
b_0 <- mh - b_1*mg
```

We write the equation of the regression line as:
$y = 0.6078*x + 105.9651$

(b) Interpret the slope and the intercept in this context.  

For each additional cm in shoulder girth, height generally increases by 0.6078 cm. When shoulder girth is 0cm, height is expected to be 106cm. It does not make sense in this context to have a shoulder girth of zero, so this just serves to adjust the height of the least squares line fit.

(c) Calculate $R^2$ of the regression line for predicting height from shoulder girth, and interpret it in the context of the application.  

```{r}
#R^2
rsqr1 <- R^2
rsqr1
```


(d) A randomly selected student from your class has a shoulder girth of 100 cm. Predict the height of this student using the model.  

```{r}
height_estimated <- b_1*100 + b_0
height_estimated
```


(e) The student from part (d) is 160 cm tall. Calculate the residual, and explain what this residual means.

```{r}
height_observed <- 160
residual_d <- height_observed - height_estimated
residual_d
```
The residual is -6.76cm. This means that the observed height value was 6.76cm less than the predicted value.

(f) A one year old has a shoulder girth of 56 cm. Would it be appropriate to use this linear model to predict the height of this child?  

No, it would not be appropriate to extrapolate for values outside of the range of the observed data that was used to generate the model.








--------------------------------------------------------------------------------

\clearpage

**Cats, Part I.** (8.26, p. 327) The following regression output is for predicting the heart weight (in g) of cats from their body weight (in kg). The coefficients are estimated using a dataset of 144 domestic cats.

\begin{center}
{
\begin{tabular}{rrrrr}
    \hline
            & Estimate  & Std. Error    & t value   & Pr($>$$|$t$|$) \\ 
    \hline
(Intercept) & -0.357    & 0.692         & -0.515    & 0.607 \\ 
body wt     & 4.034     & 0.250         & 16.119    & 0.000 \\ 
    \hline
\end{tabular} \ \\
$s = 1.452 \qquad R^2 = 64.66\% \qquad R^2_{adj} = 64.41\%$ 
}
\end{center}

\begin{center}
```{r, echo=FALSE, message=FALSE, warning=FALSE, fig.show="hold", out.width="50%", fig.height=4}
# load packages -----------------------------------------------------
library(openintro)
library(xtable)
library(MASS)
# load data ---------------------------------------------------------
data(cats)
# model heart weight vs. weight -------------------------------------
m_cats_hwt_bwt <- lm(cats$Hwt ~ cats$Bwt)
# plot heart weight vs. weight --------------------------------------
par(mar = c(3.7, 3.7, 0.5, 0.5), las = 1, mgp = c(2.5, 0.7, 0), 
    cex.lab = 1.5, cex.axis = 1.5)
plot(cats$Hwt ~ cats$Bwt, 
     xlab = "Body weight (kg)", ylab = "Heart weight (g)", 
     pch = 19, col = COL[1,2],
     xlim = c(2,4), ylim = c(5, 20.5), axes = FALSE)
axis(1, at = seq(2, 4, 0.5))
axis(2, at = seq(5, 20, 5))
box()
```
\end{center}

(a) Write out the linear model.  

$y = 4.034*x - 0.357$

(b) Interpret the intercept.  

The expected heart weight is -0.357(g) for a body weight of 0(kg). This measure does not make sense in the context of cats and onlyserves to adjust the height of the line fit.

(c) Interpret the slope.  

For every increase of 1kg of a cat's body weight, the heart weight is expected to increase by 4.034(g).

(d) Interpret $R^2$.  

With an $R^2$ of 64.66%, this means that our linear model of heart weight as a function of body weight explains 64.66% of the variability in the data.

(e) Calculate the correlation coefficient.

```{r}
Cov_hbw <- cov(cats$Bwt,cats$Hwt)
Rho <- Cov_hbw/(sd(cats$Bwt)*sd(cats$Hwt))
Rho
```




--------------------------------------------------------------------------------

\clearpage

**Rate my professor.** (8.44, p. 340) Many college courses conclude by giving students the opportunity to evaluate the course and the instructor anonymously. However, the use of these student evaluations as an indicator of course quality and teaching effectiveness is often criticized because these measures may reflect the influence of non-teaching related characteristics, such as the physical appearance of the instructor. Researchers at University of Texas, Austin collected data on teaching evaluation score (higher score means better) and standardized beauty score (a score of 0 means average, negative score means below average, and a positive score means above average) for a sample of 463 professors. The scatterplot below shows the relationship between these variables, and also provided is a regression output for predicting teaching evaluation score from beauty score.

\begin{center}
\begin{tabular}{rrrrr}
    \hline
            & Estimate  & Std. Error    & t value   & Pr($>$$|$t$|$) \\ 
  \hline
(Intercept) & 4.010     & 0.0255        & 	157.21  & 0.0000 \\ 
beauty      & 0.01325   & 0.0322        & 4.13      & 0.0000\vspace{0.8mm} \\ 
   \hline
\end{tabular}


```{r, echo=FALSE, message=FALSE, warning=FALSE, fig.show="hold", out.width="50%", fig.height=4}
# load packages -----------------------------------------------------
library(openintro)
# load data ---------------------------------------------------------
prof_evals_beauty <- read.csv("https://github.com/jbryer/DATA606Fall2019/raw/master/course_data/prof_evals_beauty.csv")
# rename variables for convenience ----------------------------------
beauty <- prof_evals_beauty$btystdave
eval <- prof_evals_beauty$courseevaluation
# model evaluation scores vs. beauty --------------------------------
m_eval_beauty = lm(eval ~ beauty)
# scatterplot of evaluation scores vs. beauty -----------------------
par(mar = c(3.9, 3.9, 0.5, 0.5), las = 0, mgp = c(2.7, 0.7, 0), 
    cex.lab = 1.5, cex.axis = 1.5, las = 1)
plot(eval ~ beauty, 
     xlab = "Beauty", ylab = "Teaching evaluation", 
     pch = 19, col = COL[1,2], 
     axes = FALSE)
axis(1, at = seq(-1, 2, 1))
axis(2, at = seq(2, 5, 1))
box()
```
\end{center}

(a) Given that the average standardized beauty score is -0.0883 and average teaching evaluation score is 3.9983, calculate the slope. Alternatively, the slope may be computed using just the information provided in the model summary table.  

```{r}
slope <- (3.9983 - 4.010)/(-0.883)
slope
```


(b) Do these data provide convincing evidence that the slope of the relationship between teaching evaluation and beauty is positive? Explain your reasoning.  

The last column in the stats table gives the p-value for the test for the coef as 0.00000. That's really small. Therefore, the data provides convincing evidence that the slope is positive.

(c) List the conditions required for linear regression and check if each one is satisfied for this model based on the following diagnostic plots.  

1. **Linearity** the data show a linear trend. this is supported by the qq plot: the data points reasonably fall on the line
2. **Nearly normal residuals** The data is reasonably normal: the residuals histogram is roughly centered on 0 and approximately normal shaped
3. **Constant variability** The variability of the residuals remains reasonably constant as a function of 'Beauty'
4. **Independent observations** There don't appear to be any trends in the 'Order of Data Collection' plot, so samples appear randomly ordered

```{r, echo=FALSE, message=FALSE, warning=FALSE, fig.show="hold", out.width="50%", fig.height=4}
# residuals plot ----------------------------------------------------
par(mar = c(3.9, 3.9, 0.5, 0.5), las = 0, mgp = c(2.7, 0.7, 0), 
    cex.lab = 1.5, cex.axis = 1.5, las = 1)
plot(m_eval_beauty$residuals ~ beauty, 
     xlab = "Beauty", ylab = "Residuals", 
     pch = 19, col = COL[1,2], 
     ylim = c(-1.82, 1.82), axes = FALSE)
axis(1, at = seq(-1, 2, 1))
axis(2, at = seq(-1, 1, 1))
box()
abline(h = 0, lty = 3)
# residuals histogram -----------------------------------------------
par(mar = c(3.9, 3, 0, 0), cex.lab = 1.5, cex.axis = 1.5)
hist(m_eval_beauty$residuals, 
     xlab = "Residuals", ylab = "", main = "",
     col = COL[1],
     xlim = c(-2,2))
# normal probability plot of residuals ------------------------------
par(mar = c(3.9, 3.9, 0.5, 0.5), mgp = c(2.7, 0.7, 0), 
    cex.lab = 1.5, cex.axis = 1.5)
qqnorm(m_eval_beauty$residuals, 
       pch = 19, col = COL[1,2],
       main = "", las = 0)
qqline(m_eval_beauty$residuals, col = COL[1])
# order of residuals ---------------------------------------------===
par(mar = c(3.9, 3.9, 0.5, 0.5), mgp = c(2.7, 0.7, 0), 
    cex.lab = 1.5, cex.axis = 1.5)
plot(m_eval_beauty$residuals, 
     xlab = "Order of data collection", ylab = "Residuals", main = "",
     pch = 19, col = COL[1,2],
     ylim = c(-1.82, 1.82), axes = FALSE)
axis(1)
axis(2, at = seq(-1, 1, 1))
box()
abline(h = 0, lty = 3)
```





