---
title: "Chapter 5 - Foundations for Inference"
author: "Bonnie Cooper"
output:
    pdf_document:
        extra_dependencies: ["geometry", "multicol", "multirow"]
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

**Heights of adults.** (7.7, p. 260) Researchers studying anthropometry collected body girth measurements and skeletal diameter measurements, as well as age, weight, height and gender, for 507 physically active individuals. The histogram below shows the sample distribution of heights in centimeters.

```{r, echo=FALSE, message=FALSE, warning=FALSE, fig.width=10, fig.height=3}
library(openintro)
data(bdims)
par(mar=c(3.7,2.5,0.5,0.5), las=1, mgp=c(2.5,0.7,0), cex.lab = 1.5)
histPlot(bdims$hgt, col = COL[1], xlab = "Height", ylab = "")
```

(a) What is the point estimate for the average height of active individuals? What about the median?

A point estimate for the average height is the mean of the sample: 
```{r}
cat( 'The point estimate for ave. height =', mean( bdims$hgt ) )
cat( '\nThe median height =', median( bdims$hgt ) )
```


(b) What is the point estimate for the standard deviation of the heights of active individuals? What about the IQR?

```{r}
cat( 'The point estimate for Standard Deviation of height =', sd( bdims$hgt ) )
cat( '\nThe IQR =', IQR( bdims$hgt ) )
```

(c) Is a person who is 1m 80cm (180 cm) tall considered unusually tall? And is a person who is 1m 55cm (155cm) considered unusually short? Explain your reasoning.

```{r}
summary( bdims$hgt )
```

180cm is greater than the 3rd Quartile (177.8) and 155sm is less than the 1st Quartile (163.8) for the height distribution. Therefore, these heights both lie outside of the Inner Quartile Range & are atypical values.

(d) The researchers take another random sample of physically active individuals. Would you expect the mean and the standard deviation of this new sample to be the ones given above? Explain your reasoning.

The mean and standard deviation of the new sample would be similar to the ones given above. The measures would not be identical due to the natural variations in the data that is sampled

(e) The sample means obtained are point estimates for the mean height of all active individuals, if the sample of individuals is equivalent to a simple random sample. What measure do we use to quantify the variability of such an estimate (Hint: recall that $SD_x = \frac{\sigma}{\sqrt{n}}$)? Compute this quantity using the data from the original sample under the condition that the data are a simple random sample.

The standard error describes how much an estimate will vary from between samples. Basically, it describes how precise the estimate is. It can be computed as follows:
```{r}
sERR <- sd( bdims$hgt )/(sqrt(length( bdims$hgt )))
cat( 'The Standard Error of the mean =', sERR )

#(mean( bdims$hgt )*(1-mean(bdims$hgt)))/length(bdims$hgt)
```





--------------------------------------------------------------------------------

\clearpage

**Thanksgiving spending, Part I.** The 2009 holiday retail season, which kicked off on November 27, 2009 (the day after Thanksgiving), had been marked by somewhat lower self-reported consumer spending than was seen during the comparable period in 2008. To get an estimate of consumer spending, 436 randomly sampled American adults were surveyed. Daily consumer spending for the six-day period after Thanksgiving, spanning the Black Friday weekend and Cyber Monday, averaged $84.71. A 95% confidence interval based on this sample is ($80.31, $89.11). Determine whether the following statements are true or false, and explain your reasoning.

```{r, echo=FALSE, message=FALSE, warning=FALSE, fig.width=10, fig.height=3}
library(openintro)
data(tgSpending)
par(mar=c(3.7,2.2,0.5,0.5), las=1, mgp=c(2.5,0.7,0), cex.lab = 1.5)
histPlot(tgSpending$spending, col = COL[1], xlab = "Spending", ylab = "")
```

(a) We are 95% confident that the average spending of these 436 American adults is between $80.31 and $89.11.  
T. If we took 100 random samples, we would expect that 95% would be within the confidence interval.<br>
(b) This confidence interval is not valid since the distribution of spending in the sample is right skewed.  
F. We can approximate a normal distribution if the samples are independent and the success-failure condition is met<br>
(c) 95% of random samples have a sample mean between $80.31 and $89.11.  
T. approximately 95% of the sample should be within the confidence limit. there may be slight deviations from 95% due to natural variations in the data.<br>
(d) We are 95% confident that the average spending of all American adults is between $80.31 and $89.11.  
F. We confident that 95% of American adults hav e an average spending between the confidence range of our sample<br>
(e) A 90% confidence interval would be narrower than the 95% confidence interval since we don't need to be as sure about our estimate.  
T. The Z-score for 90 < X-score for 95, therefore the confidence interval will span a shortr range as measured from the mean.<br>
(f) In order to decrease the margin of error of a 95% confidence interval to a third of what it is now, we would need to use a sample 3 times larger.  
F. The margin of error reduces with the square root of the sample size. So, to cut the margin by a 3rd, you would need 9x as big of a sample. <br>
(g) The margin of error is 4.4.
T:
```{r}
ME <- 1.96 * sd(tgSpending$spending)/sqrt(length(tgSpending$spending))
ME
```






--------------------------------------------------------------------------------

\clearpage

**Gifted children, Part I.** Researchers investigating characteristics of gifted children col- lected data from schools in a large city on a random sample of thirty-six children who were identified as gifted children soon after they reached the age of four. The following histogram shows the dis- tribution of the ages (in months) at which these children first counted to 10 successfully. Also provided are some sample statistics.

```{r, echo=FALSE, message=FALSE, warning=FALSE, fig.width=10, fig.height=3}
library(openintro)
data(gifted)
par(mar=c(3.7,2.2,0.5,0.5), las=1, mgp=c(2.5,0.7,0), cex.lab = 1.5)
histPlot(gifted$count, col = COL[1], 
         xlab = "Age child first counted to 10 (in months)", ylab = "", 
         axes = FALSE)
axis(1)
axis(2, at = c(0,3,6))
```

\begin{tabular}{r | l}
n   & 36 \\
min & 21 \\
mean    & 30.69 \\
sd  & 4.31 \\
max & 39 
\end{tabular}

(a) Are conditions for inference satisfied?  

**Random:** Yes, the questions mentions that the children were drawn from a random sample
**Normal** We can treat the sampling distribution as approximately normal regardless of the shape of the parent population becuase n>=30
**Independent** We can assume independence because we are sampling < 10% of the population (this is a large city, so we can assume there are more than 3600 children) and the sample is from a simple random sample

(b) Suppose you read online that children first count to 10 successfully when they are 32 months old, on average. Perform a hypothesis test to evaluate if these data provide convincing evidence that the average age at which gifted children fist count to 10 successfully is less than the general average of 32 months. Use a significance level of 0.10.

 

```{r}

sampSE <- 4.31/sqrt(36)
confInterval <- c(30.69 - qnorm(0.05)*sampSE,30.69 + qnorm(0.05)*sampSE) 
confInterval
```

The Null value (32) is within the confidence interval, therefore we cannot reject the null hypothesis.
    
(c) Interpret the p-value in context of the hypothesis test and the data.

```{r}
#Z = (sampleMean - nullMean)/SE
zed <- (30.69-32)/sampSE
pVal <- pnorm(zed)*1.645
pVal
```

The p-value is 0.056. This means that the probability of observing the same mean as the gifted sample is 5.6% if we assume that the null hypothesis is true

(d) Calculate a 90% confidence interval for the average age at which gifted children first count to 10 successfully.

```{r}
confInterval90 <- c(30.69 - qnorm(0.05)*sampSE,30.69 + qnorm(0.05)*sampSE) 
confInterval90
```


(e) Do your results from the hypothesis test and the confidence interval agree? Explain.

Yes, the p-value (0.056) < the significance level (0.10) and the null mean is outside of the confidence limits. Both of these support rejecting the null hypothesis with a significance level of 0.1




--------------------------------------------------------------------------------

\clearpage

**Gifted children, Part II.** Exercise above describes a study on gifted children. In this study, along with variables on the children, the researchers also collected data on the mother's and father's IQ of the 36 randomly sampled gifted children. The histogram below shows the distribution of mother's IQ. Also provided are some sample statistics.

```{r, echo=FALSE, message=FALSE, warning=FALSE, fig.width=10, fig.height=3}
library(openintro)
data(gifted)
par(mar=c(3.7,2.2,0.5,0.5), las=1, mgp=c(2.5,0.7,0), cex.lab = 1.5)
histPlot(gifted$motheriq, col = COL[1], 
         xlab = "Mother's IQ", ylab = "", axes = FALSE)
axis(1)
axis(2, at = c(0,4,8,12))
```

\begin{tabular}{r | l}
n   & 36 \\
min & 101 \\
mean    & 118.2 \\
sd  & 6.5 \\
max & 131 
\end{tabular}

(a) Perform a hypothesis test to evaluate if these data provide convincing evidence that the average IQ of mothers of gifted children is different than the average IQ for the population at large, which is 100. Use a significance level of 0.10.

```{r}
sampSE <- 6.5/sqrt(36)
#Z = (sampleMean - nullMean)/samSE
zed <- (118.2-100)/sampSE
pVal <- pnorm(zed)*1.645
pVal
```


(b) Calculate a 90% confidence interval for the average IQ of mothers of gifted children.

```{r}
confIntervalIQ <- c(118.2 - qnorm(0.05)*sampSE, 118.2 + qnorm(0.05)*sampSE) 
confIntervalIQ
```

(c) Do your results from the hypothesis test and the confidence interval agree? Explain.

No. population mean (100)is outside of the confidence interval which suggests we can reject the null hypothesis. However, I get a really large p-value which suggests the two means are from the same distribution. However, I must have made a mistake, because I get a p-value>1. Can't figure out what I did wrong....



--------------------------------------------------------------------------------

\clearpage

**CLT.** Define the term "sampling distribution" of the mean, and describe how the shape, center, and spread of the sampling distribution of the mean change as sample size increases.


The sampling distribution of the mean describes the normal distribution of means calculated from multiple random samplings of a distribution. Regardless of the shape of the distribution from which the means are calculated, the shape of the sampling distribution of the mean is normal. The more mean samples that are collected, the more 'normal' & symmetric the distribution. Furthermore, as sample size increases, the center of the sampling distribution of the mean approaches the true mean of the original distribution. Also, as the sample size increases, the spread of the sampling distribution of the mean gets narrower.


--------------------------------------------------------------------------------

\clearpage

**CFLBs.** A manufacturer of compact fluorescent light bulbs advertises that the distribution of the lifespans of these light bulbs is nearly normal with a mean of 9,000 hours and a standard deviation of 1,000 hours.


```{r}
x = rnorm(50000,9000,1000)
plot(density(x))
```

(a) What is the probability that a randomly chosen light bulb lasts more than 10,500 hours?

```{r}
prob <- 1 - pnorm( (10500-9000)/1000 )
prob
```


(b) Describe the distribution of the mean lifespan of 15 light bulbs.

The distribution of the mean lifespan is normal.

(c) What is the probability that the mean lifespan of 15 randomly chosen light bulbs is more than 10,500 hours?

```{r}
prob^15
```

(d) Sketch the two distributions (population and sampling) on the same scale.

```{r}
mean_sim <- 9000
std_sim <- 1000

lcb <- ((mean_sim - (3 * std_sim)) - 5)
ucb <- (((2 * mean_sim) + (3 * (2 * std_sim))) + 5)

u <- seq(from = lcb,
         to = ucb,
         length.out = 1e+5)
v1 <- dnorm(x = u,
            mean = mean_sim,
            sd = std_sim)
v2 <- dnorm(x = u,
            mean = (10500),
            sd = (std_sim))

matplot(x = u,
        y = cbind(v1, v2),
        type = "l",
        lty = 1,
        col = c("red", "blue"),
        xlim = c( 6000, 14000),
        xlab = "values",
        ylab = "densities",
        main = "overlap")
legend(x = "topright",
       legend = paste("Distbn.", 1:2),
       col = c("red", "blue"),
       lty = 1)

```

(e) Could you estimate the probabilities from parts (a) and (c) if the lifespans of light bulbs had a skewed distribution?

No, because skewed data affects the mean and standard deviation.






--------------------------------------------------------------------------------

\clearpage

**Same observation, different sample size.** Suppose you conduct a hypothesis test based on a sample where the sample size is n = 50, and arrive at a p-value of 0.08. You then refer back to your notes and discover that you made a careless mistake, the sample size should have been n = 500. Will your p-value increase, decrease, or stay the same? Explain.

The P-value will become smaller with a larger sample size. This is because, as the sample size increases, the uncertainty about the approximation of the mean decreases.




