---
title: "Chapter 6 - Inference for Categorical Data"
author: ""
output:
    pdf_document:
        extra_dependencies: ["geometry", "multicol", "multirow", "xcolor"]
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

**2010 Healthcare Law.** (6.48, p. 248) On June 28, 2012 the U.S. Supreme Court upheld the much debated 2010 healthcare law, declaring it constitutional. A Gallup poll released the day after this decision indicates that 46% of 1,012 Americans agree with this decision. At a 95% confidence level, this sample has a 3% margin of error. Based on this information, determine if the following statements are true or false, and explain your reasoning.

(a) We are 95% confident that between 43% and 49% of Americans in this sample support the decision of the U.S. Supreme Court on the 2010 healthcare law.

**False** 46% of American in the sample indicated their support for the Supreme Court's decision. There measurement is a characteristic of the sample.  

(b) We are 95% confident that between 43% and 49% of Americans support the decision of the U.S. Supreme Court on the 2010 healthcare law.  

**True** In this statement the sample statistic is used as a point estimate to project onto the American Population as a whole and is expected to lie within the confidence interval of the point statistic which is equal to the estimate (46) +/- the margin of error (3).

(c) If we considered many random samples of 1,012 Americans, and we calculated the sample proportions of those who support the decision of the U.S. Supreme Court, 95% of those sample proportions will be between 43% and 49%.

**True** This is another way of saying that we expect 95% of the new random samples to contain the first sample's point estimate (46) witin their confidence interval.

(d) The margin of error at a 90% confidence level would be higher than 3%.

**False** As the confidence level decreases, the margin of error decreases



--------------------------------------------------------------------------------

\clearpage

**Legalization of marijuana, Part I.** (6.10, p. 216) The 2010 General Social Survey asked 1,259 US residents: "Do you think the use of marijuana should be made legal, or not" 48% of the respondents said it should be made legal.

(a) Is 48% a sample statistic or a population parameter? Explain.

48% is a sample statistic, because it describes the response of the 1259 survey paticipants and not the general population.  

(b) Construct a 95% confidence interval for the proportion of US residents who think marijuana should be made legal, and interpret it in the context of the data.

```{r}
n <- 1259 #num samples
p <- 0.48 #point estimate
Zsc <- 1.96 #95% confidence
MoErr <- Zsc * sqrt( p*(1-p)/n ) #margin of error
confINT <- c( p + MoErr, p - MoErr ) #confidence int = p +/- MoErr
confINT
```

We are 95% confident that the proportion of US residents who think marjuana should be made legal is between 45.2% and 50.8%.

(c) A critic points out that this 95% confidence interval is only accurate if the statistic follows a normal distribution, or if the normal model is a good approximation. Is this true for these data? Explain.

We can check to see is a sampling distribution is nearly normal if it meets the following conditions:  

* Check that the sample's observations are independent: 
We can safely assume that this survey is a simple random sample of American adults; this means the samples are independent.
* Check that the success-failure conditions is met:
```{r}
SF1 <- n*p
SF1
SF2 <- n*(1-p)
SF2
```
Both of these values are greater than 10.

All conditions were met, therefore the normal model is an appropriate approximation.

(d) A news piece on this survey's findings states, "Majority of Americans think marijuana should be legalized." Based on your confidence interval, is this news piece's statement justified?

No, because the interval does not lie entirely above 50%. In fact, the majority of the interval is below 50%.





--------------------------------------------------------------------------------

\clearpage

**Legalize Marijuana, Part II.** (6.16, p. 216) As discussed in Exercise above, the 2010 General Social Survey reported a sample where about 48% of US residents thought marijuana should be made legal. If we wanted to limit the margin of error of a 95% confidence interval to 2%, about how many Americans would we need to survey?

```{r}
p <- 0.48 #point estimate
Zsc <- 1.96 #95% confidence

num <- Zsc^2 * p*(1-p)/ 0.02^2 
num
```
We would need 2398 Americans to participate in the survey





--------------------------------------------------------------------------------

\clearpage

**Sleep deprivation, CA vs. OR, Part I.** (6.22, p. 226) According to a report on sleep deprivation by the Centers for Disease Control and Prevention, the proportion of California residents who reported insuffient rest or sleep during each of the preceding 30 days is 8.0%, while this proportion is 8.8% for Oregon residents. These data are based on simple random samples of 11,545 California and 4,691 Oregon residents. Calculate a 95% confidence interval for the difference between the proportions of Californians and Oregonians who are sleep deprived and interpret it in context of the data.

```{r}
pCal <- 0.08
pOr <- 0.088
nCal <- 11545
nOr <- 4691
Zsc <- 1.96 #95% confidence
SE <- sqrt( pCal*(1-pCal)/nCal + pOr*(1-pOr)/nOr )
CI <- c(  ( pOr - pCal ) - Zsc*SE, ( pOr - pCal ) + Zsc*SE )
CI
```

We are 95% confident that California and Oregon have a difference of -0.0015 and 0.017 percentage. However, since 0% is contained in the interval, we do not have enough information to conclude that there is a difference in proportions.




--------------------------------------------------------------------------------

\clearpage

**Barking deer.** (6.34, p. 239) Microhabitat factors associated with forage and bed sites of barking deer in Hainan Island, China were examined from 2001 to 2002. In this region woods make up 4.8% of the land, cultivated grass plot makes up 14.7% and deciduous forests makes up 39.6%. Of the 426 sites where the deer forage, 4 were categorized as woods, 16 as cultivated grassplot, and 61 as deciduous forests. The table below summarizes these data.

\begin{center}
\begin{tabular}{c c c c c}
Woods	& Cultivated grassplot	& Deciduous forests	 & Other & Total \\
\hline 
4		& 16					& 67			     & 345	 & 426 \\
\end{tabular}
\end{center}

(a) Write the hypotheses for testing if barking deer prefer to forage in certain habitats over others.

**Null Hypothesis** Barking Deer have no foraging preference for certain habitats.
**Alternative Hypothesis** Barking Deer have a foraging preference for certain habitats over others.

(b) What type of test can we use to answer this research question?

Chi-Square test for a one-way table.

(c) Check if the assumptions and conditions required for this test
are satisfied.

**Independence** We assume that each case is independent of the other cases
**Sample Size/ distribution** Each scenario must have at least 5 expected cases: This condition is not satisfied, because there are only 4 observtions in the 'Woods' category.

(d) Do these data provide convincing evidence that barking deer pre-
fer to forage in certain habitats over others? Conduct an appro- priate hypothesis test to answer this research question.




--------------------------------------------------------------------------------

\clearpage


**Coffee and Depression.** (6.50, p. 248) Researchers conducted a study investigating the relationship between caffeinated coffee consumption and risk of depression in women. They collected data on 50,739 women free of depression symptoms at the start of the study in the year 1996, and these women were followed through 2006. The researchers used questionnaires to collect data on caffeinated coffee consumption, asked each individual about physician-diagnosed depression, and also asked about the use of antidepressants. The table below shows the distribution of incidences of depression by amount of caffeinated coffee consumption.

\small
\begin{center}
\begin{tabular}{l  l rrrrrr}
	&  \multicolumn{1}{c}{}		& \multicolumn{5}{c}{\textit{Caffeinated coffee consumption}} \\
\cline{3-7}
	&		& $\le$ 1	& 2-6	& 1	& 2-3	& $\ge$ 4	&   \\
	&		& cup/week	& cups/week	& cup/day	& cups/day	& cups/day	& Total  \\
\cline{2-8}
\textit{Clinical} & Yes	& 670 & \fbox{\textcolor{blue}{373}}	& 905	& 564	& 95 	& 2,607 \\
\textit{depression}	& No& 11,545	& 6,244	& 16,329	& 11,726	& 2,288 	& 48,132 \\
\cline{2-8}
				& Total	& 12,215	& 6,617 & 17,234	& 12,290	& 2,383 	& 50,739 \\
\cline{2-8}
\end{tabular}
\end{center}


(a) What type of test is appropriate for evaluating if there is an association between coffee intake and depression?
(b) Write the hypotheses for the test you identified in part (a).
(c) Calculate the overall proportion of women who do and do not suffer from depression.
(d) Identify the expected count for the highlighted cell, and calculate the contribution of this cell to the test statistic, i.e. ($Observed - Expected)^2 / Expected$).
(e) The test statistic is $\chi^2=20.93$. What is the p-value?
(f) What is the conclusion of the hypothesis test?
(g) One of the authors of this study was quoted on the NYTimes as saying it was "too early to recommend that women load up on extra coffee" based on just this study. Do you agree with this statement? Explain your reasoning.





