---
title: "Chapter 1 - Introduction to Data"
author: "Bonnie Cooper"
output:
    pdf_document:
        extra_dependencies: ["geometry", "multicol", "multirow"]
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

**Smoking habits of UK residents**. (1.10, p. 20) A survey was conducted to study the smoking habits of UK residents. Below is a data matrix displaying a portion of the data collected in this survey. Note that "$\pounds$" stands for British Pounds Sterling, "cig" stands for cigarettes, and "N/A" refers to a missing component of the data.

\begin{center}
\scriptsize{
\begin{tabular}{rccccccc}
\hline
	& sex 	 & age 	& marital 	& grossIncome 					     & smoke & amtWeekends	& amtWeekdays \\ 
\hline
1 	& Female & 42 	& Single 	& Under $\pounds$2,600 			     & Yes 	 & 12 cig/day   & 12 cig/day \\ 
2 	& Male	 & 44	& Single 	& $\pounds$10,400 to $\pounds$15,600 & No	 & N/A 			& N/A \\ 
3 	& Male 	 & 53 	& Married   & Above $\pounds$36,400 		     & Yes 	 & 6 cig/day 	& 6 cig/day \\ 
\vdots & \vdots & \vdots & \vdots & \vdots 				             & \vdots & \vdots 	    & \vdots \\ 
1691 & Male  & 40   & Single 	& $\pounds$2,600 to $\pounds$5,200   & Yes 	 & 8 cig/day 	& 8 cig/day \\   
\hline
\end{tabular}
}
\end{center}


(a) What does each row of the data matrix represent?  

Each row represents an observational unit, or case.  

(b) How many participants were included in the survey?  

1691

(c) Indicate whether each variable in the study is numerical or categorical. If numerical, identify as continuous or discrete. If categorical, indicate if the variable is ordinal.  

Numerical Variables: age, amtWeekends, antWeekdays. All three are discrete.
Categorical Variables: sex (nominal), marital (nominal), grossIncome (ordinal), smoke (nominal)
  
  
```{r, eval=FALSE, echo=FALSE}
# NOTE: It is not absolutely necessary to load the data to answer these 
# questions, but is here for further exploration. 
if(!file.exists('11263-Smoking_tcm86-13253.xls')) {
	download.file('https://github.com/jbryer/DATA606Fall2019/blob/master/course_data/11263-Smoking_tcm86-13253.xls?raw=true', 
				  destfile = '11263-Smoking_tcm86-13253.xls')
}
smoking <- readxl::read_excel('11263-Smoking_tcm86-13253.xls')
```





--------------------------------------------------------------------------------

\clearpage

**Cheaters, scope of inference**. (1.14, p. 29) Exercise 1.5 introduces a study where researchers studying the relationship between honesty, age, and self-control conducted an experiment on 160 children between the ages of 5 and 15[^1]. The researchers asked each child to toss a fair coin in private and to record the outcome (white or black) on a paper sheet, and said they would only reward children who report white. Half the students were explicitly told not to cheat and the others were not given any explicit instructions. Differences were observed in the cheating rates in the instruction and no instruction groups, as well as some differences across children's characteristics within each group.

(a) Identify the population of interest and the sample in this study.  

Children between the ages of 5 and 15.

(b) Comment on whether or not the results of the study can be generalized to the population, and
if the findings of the study can be used to establish causal relationships.  

The study cannot be generalized to the whole population, because it uses subjects from very targeted and very specific population of interest (children 5-15yo). 
The study establishes a causal relationship of cheating rates and self control (whether children were given instructions or not), because it is a randomized experiment that manipulates an explanatory variable (self control (instruction/no instruction)) and measures a response variable (honesty (cheating rates)).

[^1]: Alessandro Bucciol and Marco Piovesan. "Luck or cheating? A field experiment on honesty with children". In: Journal of Economic Psychology 32.1 (2011), pp. 73-78. Available at https://papers.ssrn.com/sol3/papers.cfm?abstract_id=1307694



--------------------------------------------------------------------------------

\clearpage

**Reading the paper**. (1.28, p. 31) Below are excerpts from two articles published in the NY Times:

(a) An article titled Risks: Smokers Found More Prone to Dementia states the following:

"Researchers analyzed data from 23,123 health plan members who participated in a voluntary exam and health behavior survey from 1978 to 1985, when they were 50-60 years old. 23 years later, about 25% of the group had dementia, including 1,136 with Alzheimer's disease and 416 with vascular dementia. After adjusting for other factors, the researchers concluded that pack-a- day smokers were 37% more likely than nonsmokers to develop dementia, and the risks went up with increased smoking; 44% for one to two packs a day; and twice the risk for more than two packs."

Based on this study, can we conclude that smoking causes dementia later in life? Explain your reasoning.  

The dementia study of smokers is a retrspective observational study which establishes as association of dementia with smoking. However, it does not establisha causal relationship, because this data consists of a series of observations whereas a randomized experiment would be necessary to infer causation.

(b) Another article titled The School Bully Is Sleepy states the following:

"The University of Michigan study, collected survey data from parents on each child's sleep habits and asked both parents and teachers to assess behavioral concerns. About a third of the students studied were identified by parents or teachers as having problems with disruptive behavior or bullying. The researchers found that children who had behavioral issues and those who were identified as bullies were twice as likely to have shown symptoms of sleep disorders."

A friend of yours who read the article says, "The study shows that sleep disorders lead to bullying in school children." Is this statement justified? If not, how best can you describe the conclusion that can be drawn from this study?  


The study is a prospective observational study which shows an association of sleep disorder symptoms with disruptive behavior or bullying. However, it does not demonstrate causation between the two. Therefore, saying that sleep disorders lead to bullying in school is not justified. To establish a causation, a study involving a randomized expriment would need to be implemented. Also, more objective measures of disruptive behavior would make the study more powerful, because assessments from parents and teachers relies too much on anecdotal reporting.




--------------------------------------------------------------------------------

\clearpage

**Exercise and mental health.** (1.34, p. 35) A researcher is interested in the effects of exercise on mental health and he proposes the following study: Use stratified random sampling to ensure rep- resentative proportions of 18-30, 31-40 and 41-55 year olds from the population. Next, randomly assign half the subjects from each age group to exercise twice a week, and instruct the rest not to exercise. Conduct a mental health exam at the beginning and at the end of the study, and compare the results.

(a) What type of study is this?  

Randomized Experiment with Blocking. 

(b) What are the treatment and control groups in this study?  

Treatment: exercise twice a week.
Control: no exercise. 

(c) Does this study make use of blocking? If so, what is the blocking variable?  

Yes, this study uses age for blocking into groups 18-30, 31-40 and 41-55 years of age.  

(d) Does this study make use of blinding?  

No. The subjects of the experiment can distinguish whether they are in the exerise group or not, so the study is not blind.

(e) Comment on whether or not the results of the study can be used to establish a causal rela- tionship between exercise and mental health, and indicate whether or not the conclusions can be generalized to the population at large.  

This experiment uses the random division of patients into control and treatment to establish a causal effect of exercise on mental health. However, this causal relationship should not be generalized to the population at large, because it is possible that observed differences are due to natural variations in the population. Additionally, there are many different modes of exercise that can be used at different frequencies, so there are many other experimental factors to consider befor drawing conclusions about the general population.

(f) Suppose you are given the task of determining if this proposed study should get funding. Would you have any reservations about the study proposal?  

Yes, I would have reservations about funding this study. The association of exercise to mental health could be studied more rigorously by improving the study in a number of ways (e.g. different frequncies of exercise per week. different intensities of exercise. etc.)


