---
title: "Chapter 2 - Summarizing Data"
author: ""
output:
    pdf_document:
        extra_dependencies: ["geometry", "multicol", "multirow"]
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

**Stats scores**. (2.33, p. 78) Below are the final exam scores of twenty introductory statistics students. 

57, 66, 69, 71, 72, 73, 74, 77, 78, 78, 79, 79, 81, 81, 82, 83, 83, 88, 89, 94

Create a box plot of the distribution of these scores. The five number summary provided below may be useful.

\begin{center}
\renewcommand\arraystretch{1.5}
\begin{tabular}{ccccc}
Min & Q1    & Q2 (Median)   & Q3    & Max \\
\hline
57  & 72.5  & 78.5          & 82.5  & 94 \\
\end{tabular}
\end{center}

```{r, echo = FALSE}
scores <- c(57, 66, 69, 71, 72, 73, 74, 77, 78, 78, 79, 79, 81, 81, 82, 83, 83, 88, 89, 94)

```

**My Solution**:
Using the boxplot() function to create a boxplot of the final exam scores

```{r}
scoresBoxplot <- boxplot( scores,
         main = 'Distribution of Student Final Exam Scores',
         xlab = 'Student Group',
         ylab = 'Exam Score' )

cat( 'The scoresBoxplot has a lower whisker = ', scoresBoxplot$stats[1],
     '\na lower hinge (1st quartile) = ', scoresBoxplot$stats[2], 
     '\na median value = ', scoresBoxplot$stats[3], 
     '\nan upper hinge (3rd quartile) = ', scoresBoxplot$stats[4],
     '\nand an upper whisker = ', scoresBoxplot$stats[5],
     '\nAdditionally, there is one outlier value, ', scoresBoxplot$out)
```



--------------------------------------------------------------------------------

\clearpage

**Mix-and-match**. (2.10, p. 57) Describe the distribution in the histograms below and match them to the box plots.

```{r, echo=FALSE, message=FALSE, warning=FALSE, fig.width=10, fig.height=3}
library(openintro)
set.seed(7365)
sym = rnorm(1000, mean = 60, sd = 3)
uni = runif(1000, min = 0, max = 100)
rs = rgamma(1000, shape = 3, rate = 2)
par(mar=c(4, 3.6, 0, 0), las = 1, mgp = c(2.7, 0.7, 0), 
    mfrow = c(1,6), 
    cex.lab = 1.5, cex.axis = 1.5)
histPlot(sym, xlab = "(a)", ylab = "", col = COL[1], axes = FALSE)
axis(1, seq(50,70,10))
histPlot(uni, xlab = "(b)", ylab = "", col = COL[1], axes = FALSE)
axis(1, seq(0,100,50))
histPlot(rs, xlab = "(c)", ylab = "", col = COL[1], axes = FALSE)
axis(1, seq(0,6,2))
boxPlot(rs, xlab = "(1)", ylab = "", col = COL[1,3])
boxPlot(sym, xlab = "(2)", ylab = "", col = COL[1,3])
boxPlot(uni, xlab = "(3)", ylab = "", col = COL[1,3])
```
<br><br>  

**My Solution**:  

* Histogram 'a' is best described as a symmetric data set and corresponds to boxplot 2.
* Histrogram 'b' is a uniform or flat distribution; this data set corresponds to boxplot 3.
* Histogram 'c' is right skewed and corresponds to boxplot 1.




--------------------------------------------------------------------------------

\clearpage

**Distributions and appropriate statistics, Part II**. (2.16, p. 59) For each of the following, state whether you expect the distribution to be symmetric, right skewed, or left skewed. Also specify whether the mean or median would best represent a typical observation in the data, and whether the variability of observations would be best represented using the standard deviation or IQR. Explain your reasoning.

(a) Housing prices in a country where 25% of the houses cost below $350,000, 50% of the houses cost below $450,000, 75% of the houses cost below $1,000,000 and there are a meaningful number of houses that cost more than $6,000,000.
(b) Housing prices in a country where 25% of the houses cost below $300,000, 50% of the houses cost below $600,000, 75% of the houses cost below $900,000 and very few houses that cost more than $1,200,000.
(c) Number of alcoholic drinks consumed by college students in a given week. Assume that most of these students don't drink since they are under 21 years old, and only a few drink excessively.
(d) Annual salaries of the employees at a Fortune 500 company where only a few high level
executives earn much higher salaries than the all other employees.


**My Response**.  

(a) Right-skewed: this distribution would be expected to have a mean value that is skewed to the right of the media by all the values that lie far above the 3rd quartile (>1M). The median would be the best representation of a typical data observation and the IQR would best represent the variability of the data, because these measures are more robust to the effect of the outlier data points.  

(b) Symmetric, because the 1st and 3rd quartiles are equidistant about the median and there are very few values expected to be > 1.5 the IQR. For this distribution, because the effect of outliers is expected to be minimal, the mean would be the best representation of a typical data observations and the standard deviation would be best to capture the spread of the data.  

(c) Right-skewed: because most of the student population is under 21, the majority of the reported drink frequencies will be 0, a substantial proportion of students will report drinking 1-several drinks a week. However, there are be relatively few students whom report drinking many beers/week. (This might resemble a poisson distribution?). The students who report drining many beers a weekwill skew the mean od the data to the right of the median. Because of this skew to the data distribution, median would be the best to estiate a typical data observation and the IQR would best represent the variability of the data  (these measures are more robust to the effects of the outliers).  

(d) Left-skewed: I would expect this distribution to be left skewed, but this might only serve to reveal how naive I am to corporate America. I assume there will be no employees working for free, many people working low-medum wage supportive roles, a large proportion making above the U.S. median wage, and then a sharp fall-off in numbers with the relatively few very high-earning CEOs. As with the other skewed data distributions, the median would be the best to represent a typical data observation and the IQR would best represent the variability of the data (these measures are more robust to the effect of the outliers).




--------------------------------------------------------------------------------

\clearpage

**Heart transplants.** (2.26, p. 76) The Stanford University Heart Transplant Study was conducted to determine whether an  experimental heart transplant program increased lifespan. Each patient entering the program was designated an official heart transplant candidate, meaning that he was gravely ill and would most likely benefit from a new heart. Some patients got a transplant and some did not. The variable *transplant* indicates which group the patients were in; patients in the treatment group got a transplant and those in the control group did not. Of the 34 patients in the control group, 30 died. Of the 69 people in the treatment group, 45 died. Another variable called *survived* was used to indicate whether or not the patient was alive at the end of the study.

```{r, echo=FALSE, message=FALSE, warning=FALSE, fig.show="hold", out.width="50%", fig.height=3}
library(openintro)
data(heartTr)
# mosaic plot -------------------------------------------------------
par(mar = c(0, 0, 0, 0), las = 1, mgp = c(2.7, 0.9, 0))
mosaicplot(transplant ~ survived, data = heartTr, 
           main = "", xlab = "", ylab = "", color = COL[1],
           cex.axis = 1.5)

# box plot ----------------------------------------------------------
par(mar = c(2, 4.8, 0, 0), las = 1, mgp = c(3.5, 0.7, 0), 
    cex.lab = 1.5, cex.axis = 1.5)
boxPlot(heartTr$survtime, fact = heartTr$transplant, 
        ylab = "Survival Time (days)", col = COL[1,2])
```

(a) Based on the mosaic plot, is survival independent of whether or not the 
patient got a transplant? Explain your reasoning.

**My Response**.  
The mosaic plot suggests that survival is not independent of receiving a transplant, because a higher proportion of patients that received transplants survived when compared to the control group. However, the sample number is rather small, so it is unclear from this plor alone whether this is convincing evidence for an effect of treatment.
<br>

(b) What do the box plots below suggest about the efficacy (effectiveness) of the heart transplant treatment.

**My Response**.  
The box plots show the distribution of survival time in days. This plot compares the control and treatment groups side-by-side. This comparison reveals that the median survival time is greater for patients that recieved the treatments compared to the control group. Additionally, the spread of the data given by the IQR bars shows that the range of the survival times for patients that recieved transplants spans a much greater time-range than the control group.
<br>

(c) What proportion of patients in the treatment group and what proportion of 
patients in the control group died?

**My Response**.
```{r}
#make an R data.frame of the data
transplantResults <- data.frame( "group"=c('control','treatment'),
                                 "total"=c(34,69),
                                 "died"=c(30,45))
#add a column that calculates the proportion of each group that died
transplantResults$propdied <- transplantResults$died / 
  transplantResults$total
#format some output text
cat( 'Proportion of patients in treatment group that died = ', 
     transplantResults$propdied[2], 
     '\nProportion of patients in control group that died = ',
     transplantResults$propdied[1])
```


(d) One approach for investigating whether or not the treatment is effective 
is to use a randomization technique.

i. What are the claims being tested?

**My Response**.  
the randomization technique tests the two competing claims:

* Independence model:  that treatment and survival outcome are independent.
* Alternative: that treatment and survival outcome are NOT independent.
<br>

ii. The paragraph below describes the set up for such approach, if we were 
to do it without using statistical software. Fill in the blanks with a number 
or phrase, whichever is appropriate.


```{r}
transplantResults$survived <- transplantResults$total - 
  transplantResults$died
transplantResTotals <- colSums(transplantResults[,c(2,3,4,5)])
transplantResTotals
```


> We write *alive* on ___28___ cards representing patients who were 
alive at the end of the study, and *dead* on ___75___ cards 
representing patients who were not. Then, we shuffle these cards and split them 
into two groups: one group of size ___69___ representing treatment, and 
another group of size ___34___ representing control. We calculate the 
difference between the proportion of *dead* cards in the treatment and 
control groups (treatment - control) and record this value. We repeat this 100 
times to build a distribution centered at ___0___. Lastly, we calculate 
the fraction of simulations where the simulated differences in proportions are 
___0.88-0.65=0.23___ ___(The proportion with >= the observed difference in proportions)___. If this fraction is low, we conclude that it is unlikely to 
have observed such an outcome by chance and that the null hypothesis should 
be rejected in favor of the alternative.

iii. What do the simulation results shown below suggest about the effectiveness 
of the transplant program?

```{r fig.height=3, fig.show="hold", message=FALSE, warning=FALSE, include=FALSE, out.width="50%", results='hide'}
# randomization -----------------------------------------------------
diffs <- DATA606::inference(heartTr$survived, heartTr$transplant, 
                  success = "dead", order = c("treatment","control"), 
                  est = "proportion", type = "ht", method = "simulation", 
                  nsim = 100, null = 0, alternative = "twosided", simdist = TRUE,
                  seed = 95632)
```

\begin{center}
```{r, results='hide', echo=FALSE, message=FALSE, warning=FALSE, fig.show="hold", out.width="75%", fig.height=3}
# plot randomization distribution -----------------------------------
par(mar = c(3.6, 2.2, 1, 1), las = 1, mgp = c(2.5, 0.7, 0), 
    cex.axis = 1.5, cex.lab = 1.5)
values  <- table(diffs)
plot(diffs, type = "n", xlim = c(-0.25, 0.25), 
     xlab = "simulated differences in proportions", 
     ylab = "", axes = FALSE, ylim = c(0, max(values)))
axis(1, at = seq(-0.25, 0.25, 0.05), 
     labels = c(-0.25, NA,-0.15, NA,-0.05, NA, 0.05, NA, 0.15, NA, 0.25))
for(i in 1:length(diffs)){
  x   <- diffs[i]
  rec <- sum(diffs == x)
  points(rep(x, rec), 1:rec, pch = 20, cex = 0.8, col = COL[1])
}
```
\end{center}

**My Response**.  
The observed difference in proportions: 0.8823529-0.6521739 = 0.230179  
The histogram of the simulated differences in proportions do not appear to have a single value greater than 0.23. Therefore, we conclude that the evidence is sufficiently strong to reject the null hypothesis and assert that there is convincing evidence that the heart transplant treatment is an effect of treatment compared to the control group.



