---
title: "Chapter 3 - Probability"
author: "Bonnie Cooper"
output:
    pdf_document:
        extra_dependencies: ["geometry", "multicol", "multirow"]
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(error = TRUE)
```

**Dice rolls.** (3.6, p. 92) If you roll a pair of fair dice, what is the probability of

(a) getting a sum of 1?
(b) getting a sum of 5?
(c) getting a sum of 12?


Let start by simulating all possible die rolls:

```{r}
library( cocoreg )
dice1 <- c( 1, 2, 3, 4, 5, 6 ) #all possible outcomes for dice 1
dice2 <- c( 1, 2, 3, 4, 5, 6 ) #all possible outcomes for dice 2
diceOutcomes <- matrix( rep( 0, length( dice1 )*length( dice2 ) ), 
                    nrow = length( dice1 ) ) #alocate space for outcomes
diceAdd <- list() #initialize a list for addition in our for loop
diceAdd[[1]] <- dice1 #add dice1 to the diceAdd list
for( ii in 1:length( dice2 ) ){  #for each value in dice2
  diceAdd[[2]] <- dice2  #add dice2 to the diceAdd list
  diceOutcomes[ ,ii ] <- do.call( '+', diceAdd ) #do.call() to perform element wise addition
  dice2 <- cshift( dice1, ii ) #circular shift dice1 to add in next loop
}
diceOutcomes
```
<br> 

The matrix above holds a value for each possible outcome. From this, we can find the information we need to calculate the probabilities we are interested in

```{r}
outcomesFrequencyTable <- table( diceOutcomes )
numPossibleOutcomes <- length( diceOutcomes )
cat( 'There are', numPossibleOutcomes, 'possible outcomes\n\n' )
cat( 'This table gives the frequency (count) of the outcome values:\n')
outcomesFrequencyTable
```

(a) What is the probability of getting a sum of 1?
```{r }
test <- '0'
result <- outcomesFrequencyTable[[test]]/numPossibleOutcomes
cat( 'the probability of getting a sum of', test, 'is', result)
```
We get an error because there is no value, 1, in 'diceOutcomes' and this makes perfect sense: We rolled 2 die where the minimum value is 1, so the minimum possible sum is 2. The probability of rolling 2 die and summing their values to 1 is zero, P(1) = 0


(b) getting a sum of 5?
```{r}
test <- '5'
result <- outcomesFrequencyTable[[test]]/numPossibleOutcomes
cat( 'P(', test, ') is', result)
```



(c) getting a sum of 12?
```{r}
test <- '12'
result <- outcomesFrequencyTable[[test]]/numPossibleOutcomes
cat( 'P(', test, ') is', result)
```

--------------------------------------------------------------------------------

\clearpage

**Poverty and language**. (3.8, p. 93) The American Community Survey is an ongoing survey that provides data every year to give communities the current information they need to plan investments and services. The 2010 American Community Survey estimates that 14.6% of Americans live below the poverty line, 20.7% speak a language other than English (foreign language) at home, and 4.2% fall into both categories.

(a) Are living below the poverty line and speaking a foreign language at home disjoint?
 

4.2% Americans both speak another language at home and live below the poverty line, therefore these two variables are not disjoint, because disjoint variables by cannot both happen simultaneously.  
<br>

(b) Draw a Venn diagram summarizing the variables and their associated probabilities.


```{r, message=FALSE}

#load the VennDiagram demo
library( VennDiagram )

# Plot our Venn Diagram
grid.newpage()
invisible(
  draw.pairwise.venn(area1 = 14.6, 
                   area2 = 20.7, 
                   cross.area = 4.2, 
                   category = c("Living Below\nPoverty Line\n14.6%  of Americans",
                              "Second Language\nSpoken at Home\n20.7% of Americans"),
                   lty = rep("blank", 2), 
                   fill = c("light blue", "pink"), 
                   alpha = rep(0.5, 2), 
                   cat.pos = c(0, 0), 
                   cat.dist = rep(0.025, 2)))
```


(c) What percent of Americans live below the poverty line and only speak English at home?  
  
10.4% of Americans live below the poverty line and only speak English at home  
<br>

(d) What percent of Americans live below the poverty line or speak a foreign language at home?  
 
16.5 + 10.4= 26.9% of Americans either live below the poverty line or speak a foreign language.  
<br> 

(e) What percent of Americans live above the poverty line and only speak English at home?  
  
100 - 16.5 - 4.2 - 10.4 = 68.9% of Americans live above the poverty line and only speak English at home.  
<br>

(f) Is the event that someone lives below the poverty line independent of the event that the person speaks a foreign language at home?  
  
The Multiplication Rule for Independent Processes tells us that for 'Living Above the Poverty Line' (LAPL) and 'Second Language Spoken at Home' (SLSH) to be independent:   
P($LAPL \cap SLSH$) = P(LAPL) x P(SLSH)  
However, 0.042 $\neq$ 0.207 x 0.146  
0.207 x 0.146 = 0.0302
Therefore, if someone lives below the poverty line, this is not independent of the same person speaking a second language at home.  
<br>




--------------------------------------------------------------------------------

\clearpage

**Assortative mating**. (3.18, p. 111) Assortative mating is a nonrandom mating pattern where individuals with similar genotypes and/or phenotypes mate with one another more frequently than what would be expected under a random mating pattern. Researchers studying this topic collected data on eye colors of 204 Scandinavian men and their female partners. The table below summarizes the results. For simplicity, we only include heterosexual relationships in this exercise.

\begin{center}
\begin{tabular}{ll  ccc c}
                                        &           & \multicolumn{3}{c}{\textit{Partner (female)}} \\
\cline{3-5}
                                        &           & Blue  & Brown     & Green     & Total \\
\cline{2-6}
                                        & Blue      & 78    & 23        & 13        & 114 \\
\multirow{2}{*}{\textit{Self (male)}}   & Brown     & 19    & 23        & 12        & 54 \\
                                        & Green     & 11    & 9         & 16        & 36 \\
\cline{2-6}
                                        & Total     & 108   & 55        & 41        & 204
\end{tabular}
\end{center}


Before answering the questions, let's calculate the relative frequencies from the raw numbers.
```{r, message=FALSE}
library(dplyr)

eyeColor_mat <- matrix( 
  c( 78, 23, 13, 114, 19, 23, 12, 54, 11, 9, 16, 36, 108, 55, 41, 204 ),
  nrow = 4,
  ncol = 4,
  byrow = TRUE )
eyeColor_mat <- eyeColor_mat/204
eyeColor_DF <- as.data.frame( t( eyeColor_mat ) )
colnames( eyeColor_DF ) <- c( 'F_Blue', 'F_Brown', 'F_Green', 'F_Total' )
eyeColor_DF <- eyeColor_DF %>% mutate_if( is.numeric, round, digits = 3 )
rownames( eyeColor_DF ) <- c( 'M_Blue', 'M_Brown', 'M_Green', 'M_Total' )
eyeColor_DF
```



(a) What is the probability that a randomly chosen male respondent or his partner has blue eyes?    
  
P( $M_Blue \cup F_Blue$ ) = P( M_Blue ) + P( F_Blue ) - P( $M_Blue \cap F_Blue$ )  
P( $M_Blue \cup F_Blue$ ) = 0.529 + 0.559 - 0.382  
P( $M_Blue \cup F_Blue$ ) = 0.706  
<br>

(b) What is the probability that a randomly chosen male respondent with blue eyes has a partner with blue eyes?  
  
P( $M_Blue \cap F_Blue$ ) = 0.382
<br>

(c) What is the probability that a randomly chosen male respondent with brown eyes has a partner with blue eyes?  
  
P( $M_Brown \cap F_Blue$ ) = 0.113
<br>

What about the probability of a randomly chosen male respondent with green eyes having a partner with blue eyes?  
 
P( $M_Green \cap F_Blue$ ) = 0.064
<br>

(e) Does it appear that the eye colors of male respondents and their partners are independent? Explain your reasoning.  
  
We can use the  Multiplication Rule for Independent Processes to test whether the eye colors of male responsents and their partners are independent. 
If they are independent, then the P( $M_Blue \cap F_Blue$ ) should equal the product of P( M_Blue ) x P( F_Blue ).  
However,  
0.382 $\neq$ 0.559*0.529  
0.559*0.529 = 0.296  
Therefore the eye color of male respondents and their partners is not independent.




--------------------------------------------------------------------------------

\clearpage

**Books on a bookshelf**. (3.26, p. 114) The table below shows the distribution of books on a bookcase
based on whether they are nonfiction or fiction and hardcover or paperback.

\begin{center}
\begin{tabular}{ll  cc c}
                                &           & \multicolumn{2}{c}{\textit{Format}} \\
\cline{3-4}
                                &           & Hardcover     & Paperback     & Total \\
\cline{2-5}
\multirow{2}{*}{\textit{Type}}  & Fiction   & 13            & 59            & 72 \\
                                & Nonfiction& 15            & 8             & 23 \\
\cline{2-5}
                                & Total     & 28            & 67            & 95 \\
\cline{2-5}
\end{tabular}
\end{center}

(a) Find the probability of drawing a hardcover book first then a paperback fiction book second when drawing without replacement.  
28/95 x 59/94 %~~% 0.185 
<br>

(b) Determine the probability of drawing a fiction book first and then a hardcover book second, when drawing without replacement.  
72/95 x 28/94 %~~% 0.226  
<br>

(c) Calculate the probability of the scenario in part (b), except this time complete the calculations under the scenario where the first book is placed back on the bookcase before randomly drawing the second book.
72/95 x 28/95 %~~% 0.223  
<br>

(d) The final answers to parts (b) and (c) are very similar. Explain why this is the case.

The answers are very similar, because removing one book has a small effect on the probability of what book might be drawn next. The effect on probability would be greater if their were a longer series of books checked out to compare with and without replacement or if the total number of books were significantly reduced.



--------------------------------------------------------------------------------

\clearpage

**Baggage fees**. (3.34, p. 124) An airline charges the following baggage fees: $25 for the first bag and $35 for the second. Suppose 54% of passengers have no checked luggage, 34% have one piece of checked luggage and 12% have two pieces. We suppose a negligible portion of people check more than two bags.

(a) Build a probability model, compute the average revenue per passenger, and compute the corresponding standard deviation.

```{r}
# Probability model given as a linear combination of variables
#E_Revenue = 0*( No_Bag ) + 25*( First_Bag ) + 35*( Second_Bag ) ->
E_Revenue <- 25*0.34 + 35*0.12
cat( 'The Expected Revenue (average revenue/passenger) =', E_Revenue,'\n' )

Var_First_Bag <- ( 25-E_Revenue )^2*0.34
Var_Second_Bag <- ( 35-E_Revenue )^2*0.12
Variance <- 25^2*Var_First_Bag + 35^2*Var_Second_Bag
STDev <- sqrt( Variance )
cat( 'The corresponding Standard Deviation =', STDev )
```


(b) About how much revenue should the airline expect for a flight of 120 passengers? With what standard deviation? Note any assumptions you make and if you think they are justified.

For this linear combination model, we assume that the random variables (First_Bag & Second_Bag) are independent.
The airline should expect to make approximately 12.7x120 = $1524
The Standard Deviation should be 324.4237




--------------------------------------------------------------------------------

\clearpage

**Income and gender**. (3.38, p. 128) The relative frequency table below displays the distribution of annual total personal income (in 2009 inflation-adjusted dollars) for a representative sample of 96,420,486 Americans. These data come from the American Community Survey for 2005-2009. This sample is comprised of 59% males and 41% females.

\begin{center}
\begin{tabular}{lr}
  \hline
\textit{Income}         & \textit{Total} \\
  \hline
\$1 to \$9,999 or loss  & 2.2\% \\
\$10,000 to \$14,999    & 4.7\% \\
\$15,000 to \$24,999    & 15.8\% \\
\$25,000 to \$34,999    & 18.3\% \\
\$35,000 to \$49,999    & 21.2\% \\
\$50,000 to \$64,999    & 13.9\% \\
\$65,000 to \$74,999    & 5.8\% \\
\$75,000 to \$99,999    & 8.4\% \\
\$100,000 or more       & 9.7\% \\
   \hline
\end{tabular}
\end{center}

(a) Describe the distribution of total personal income.
```{r}
df <- data.frame( 
  income = c( 1, 10000, 15000, 25000, 35000, 50000, 65000, 75000, 100000),
  total = c( 2.2, 4.7, 15.8, 18.3, 21.2, 13.9, 5.8, 8.4, 9.7))
library( ggplot2 )
ggplot( df, aes(x=income, y=total)) + geom_bar(stat= 'identity')
```
<br>  
The bar graph above gives a sketch of the total personal income data distribution. the distribution is that of a probability density function for a continuous variable.

(b) What is the probability that a randomly chosen US
resident makes less than $50,000 per year?

P(<$50,000) can be found by summing the probability values that span from 1 to 49,999.  
P(<$50,000) = 0.022 + 0.047 + 0.158 + 0.183 + 0.212  
P(<$50,000) = 0.622 (62.2%)   
<br>

(c) What is the probability that a randomly chosen US resident makes less than $50,000 per year and is female?
Note any assumptions you make.

If we assume that gender and income are independent and that the male:female ratio of the sample is representative of the population, then we estimate the probability to be the P(<$50,000)xP(Female) = 0.622x0.41 = 0.255 (25.5)
<br>

(d) The same data source indicates that 71.8% of females
make less than $50,000 per year. Use this value to determine whether or not the assumption you made in part (c) is valid.

We can use Bayes Theorem to invert the probability since we know that:
P(Female) = 0.41   
P(makes<$50k and Female) = 0.255  
P(Female | makes<$50k) = P(makes<$50k and Female)/P(Female)  
P(Female | makes<$50k) = 0.622  

The data source found that 71.8% of females make less than 50k/year & this is much higher than the estimate that we calculated. This tells us that out initial assumption that gender and income are independent is not correct.



