---
title: "Multiple linear regression"
output:
  pdf_document: default
  html_document:
    css: ./lab.css
    highlight: pygments
    theme: cerulean
---

## Grading the professor

Many college courses conclude by giving students the opportunity to evaluate 
the course and the instructor anonymously. However, the use of these student 
evaluations as an indicator of course quality and teaching effectiveness is 
often criticized because these measures may reflect the influence of 
non-teaching related characteristics, such as the physical appearance of the 
instructor. The article titled, "Beauty in the classroom: instructors' 
pulchritude and putative pedagogical productivity" (Hamermesh and Parker, 2005) 
found that instructors who are viewed to be better looking receive higher 
instructional ratings. (Daniel S. Hamermesh, Amy Parker, Beauty in the 
classroom: instructors pulchritude and  putative pedagogical productivity, 
*Economics of Education Review*, Volume 24, Issue 4, August 2005, Pages 369-376, 
ISSN 0272-7757, 10.1016/j.econedurev.2004.07.013. [http://www.sciencedirect.com/science/article/pii/S0272775704001165](http://www.sciencedirect.com/science/article/pii/S0272775704001165).)

In this lab we will analyze the data from this study in order to learn what goes 
into a positive professor evaluation.

## The data

The data were gathered from end of semester student evaluations for a large 
sample of professors from the University of Texas at Austin. In addition, six 
students rated the professors' physical appearance. (This is aslightly modified 
version of the original data set that was released as part of the replication 
data for *Data Analysis Using Regression and Multilevel/Hierarchical Models* 
(Gelman and Hill, 2007).) The result is a data frame where each row contains a 
different course and columns represent variables about the courses and professors.

```{r load-data, eval=TRUE}
load("more/evals.RData")
```

variable         | description
---------------- | -----------
`score`          | average professor evaluation score: (1) very unsatisfactory - (5) excellent.
`rank`           | rank of professor: teaching, tenure track, tenured.
`ethnicity`      | ethnicity of professor: not minority, minority.
`gender`         | gender of professor: female, male.
`language`       | language of school where professor received education: english or non-english.
`age`            | age of professor.
`cls_perc_eval`  | percent of students in class who completed evaluation.
`cls_did_eval`   | number of students in class who completed evaluation.
`cls_students`   | total number of students in class.
`cls_level`      | class level: lower, upper.
`cls_profs`      | number of professors teaching sections in course in sample: single, multiple.
`cls_credits`    | number of credits of class: one credit (lab, PE, etc.), multi credit.
`bty_f1lower`    | beauty rating of professor from lower level female: (1) lowest - (10) highest.
`bty_f1upper`    | beauty rating of professor from upper level female: (1) lowest - (10) highest.
`bty_f2upper`    | beauty rating of professor from second upper level female: (1) lowest - (10) highest.
`bty_m1lower`    | beauty rating of professor from lower level male: (1) lowest - (10) highest.
`bty_m1upper`    | beauty rating of professor from upper level male: (1) lowest - (10) highest.
`bty_m2upper`    | beauty rating of professor from second upper level male: (1) lowest - (10) highest.
`bty_avg`        | average beauty rating of professor.
`pic_outfit`     | outfit of professor in picture: not formal, formal.
`pic_color`      | color of professor's picture: color, black & white.

## Exploring the data

1.  Is this an observational study or an experiment? The original research
    question posed in the paper is whether beauty leads directly to the
    differences in course evaluations. Given the study design, is it possible to
    answer this question as it is phrased? If not, rephrase the question.
    
    This is an observational study, because the data was collected in a way that did not interfere with how the data arose: Professors physical appearances were rated by an independent panel of students after the fact that the data set had been collected.

2.  Describe the distribution of `score`. Is the distribution skewed? What does 
    that tell you about how students rate courses? Is this what you expected to 
    see? Why, or why not?
    
    As shown in the histogram below, `score` is left-skewed (mean is shifted to the left of the mode by the long 'tail' to the left). This tells us that typical students have a tendancy to rate courses favorably while a smaller minority give less favorable ratings. This outcome is somewhat surprizing because the students have no incentive to rank courses higher, so it is interesting (and encouraging) that the students rank classes favorably. It would be interesting to see how this distribution correlates with student grades.
    
```{r, message=F}
library(ggplot2)
ggplot(evals, aes(x=score)) + 
  geom_histogram()
```
    

3.  Excluding `score`, select two other variables and describe their relationship 
    using an appropriate visualization (scatterplot, side-by-side boxplots, or 
    mosaic plot).
    
    Based on the side-byside boxplots and the envelope of the data series histograms, the varables `bty_avg` and `bty_f1lower` have less skew and follow a more normal distribution than the `score` variable. `bty_avg` and `bty_f1lower` have very similar distribution & this is expected, because `bty_avg` includes `bty_f1lower` in it's calculation.
    
```{r, message=F}
library( tidyverse )

plotData <- evals %>%
  select("score", "bty_avg", "bty_f1lower") %>%
  pivot_longer(cols = everything(),names_to = "var", values_to = "val")

plotBox <- ggplot(plotData, aes(var, val, fill=var)) + 
  geom_boxplot() + 
  labs(title = "Side-by-side Boxplots")
plotBox

plotHist <- ggplot(plotData, aes( x=val, fill=var )) +
  geom_histogram( color="black", alpha=0.6, position = 'identity') +
    scale_fill_manual(values=c("#5233FF", "#33FF4F", "#E033FF")) +
    labs(title = 'Overlapped Histograms')
plotHist
```
    

## Simple linear regression

The fundamental phenomenon suggested by the study is that better looking teachers
are evaluated more favorably. Let's create a scatterplot to see if this appears 
to be the case:

```{r scatter-score-bty_avg}
plot(evals$score ~ evals$bty_avg)
```


Before we draw conclusions about the trend, compare the number of observations 
in the data frame with the approximate number of points on the scatterplot. 
Is anything awry?

4.  Replot the scatterplot, but this time use the function `jitter()` on the
    $y$- or the $x$-coordinate. (Use `?jitter` to learn more.) What was
    misleading about the initial scatterplot?
    
    The initial plot was misleading, because many of the datapoints were overlapping; this did not give an accurate representation of the overall distribution of the data points.
    
```{r}
plot(evals$score ~ jitter(evals$bty_avg))
```
    

5.  Let's see if the apparent trend in the plot is something more than
    natural variation. Fit a linear model called `m_bty` to predict average
    professor score by average beauty rating and add the line to your plot
    using `abline(m_bty)`. Write out the equation for the linear model and
    interpret the slope. Is average beauty score a statistically significant
    predictor? Does it appear to be a practically significant predictor?
    
    The least squares regression line for the linear model:
\[
  \hat{y} = 3.88034 + 0.06664 * bty\_avg
\]
```{r}
m_bty <- lm(score ~ bty_avg, data = evals)
summary(m_bty)

plot(evals$score ~ jitter(evals$bty_avg))
abline(m_bty)
```
    
    Average beauty score is a statistically significant predictor, because the p-value is quite small (5.083e-05). However, average beauty is not a practical sigificant predictor because the slope is very small (0.0666) indicating that rating does not change much as a function of averge beauty. Also, the R^2 is very small (3.2%) indicating that average_beauty does not describe much of the data's variance.

6.  Use residual plots to evaluate whether the conditions of least squares
    regression are reasonable. Provide plots and comments for each one (see
    the Simple Regression Lab for a reminder of how to make these).
    
    **Model Diagnostics**  
    **Linearity & Constant Variability**  
    check if the relationship is linear with a scatterplot and evaluate the variability as a function of bty_avg.
```{r}
plot( m_bty$residuals ~ jitter(evals$bty_avg))
abline( h=0, lty=3 )
```
    
    Based on the scatterplot of the residuals above, the data appear to be reasonably linear and there is no obvious hange in variability related to the value of bty_avg
    
    **Nearly Normal Residuals**
    use a histogram & qq-plot to check for normality
```{r}
hist( m_bty$residuals )
```
```{r}
qqnorm( m_bty$residuals )
qqline( m_bty$residuals )
```
    
    The histogram of residuals appears skewed, but does not show any obvious signs of outliers. Additionally, the normal probability plot follows a straight line reasonably well.
    In conclusion, the conditions for this linear model are reasonably met.

## Multiple linear regression

The data set contains several variables on the beauty score of the professor: 
individual ratings from each of the six students who were asked to score the 
physical appearance of the professors and the average of these six scores. Let's 
take a look at the relationship between one of these scores and the average 
beauty score.

```{r bty-rel}
plot(evals$bty_avg ~ evals$bty_f1lower)
cor(evals$bty_avg, evals$bty_f1lower)
```

As expected the relationship is quite strong - after all, the average score is 
calculated using the individual scores. We can actually take a look at the 
relationships between all beauty variables (columns 13 through 19) using the 
following command:

```{r bty-rels}
plot(evals[,13:19])
```

These variables are collinear (correlated), and adding more than one of these 
variables to the model would not add much value to the model. In this 
application and with these highly-correlated predictors, it is reasonable to use
the average beauty score as the single representative of these variables.

In order to see if beauty is still a significant predictor of professor score 
after we've accounted for the gender of the professor, we can add the gender 
term into the model.

```{r scatter-score-bty_avg_gender}
m_bty_gen <- lm(score ~ bty_avg + gender, data = evals)
summary(m_bty_gen)
```

7.  P-values and parameter estimates should only be trusted if the
    conditions for the regression are reasonable. Verify that the conditions
    for this model are reasonable using diagnostic plots.
    
    **Model Diagnostics** (following 9.4.3 Checking model conditions using graphs)  
    **Nearly Normal Residuals:** Check for outliers on residuals histogram
    There are no obvious sign of 'major' outliers in the distribution
```{r}
hist( m_bty_gen$residuals )
```

    **Check for constant variance:** absolute value of residuals against the fitted vlue
    There are no obvious deviations from constant variance

```{r}
plot(abs(m_bty_gen$residuals) ~ jitter(m_bty_gen$fitted.values))
```

    **Data Independence:** Residuals in Order of their data collection
    There is no structure to the data as a function of data collection, so we can assume indepenence.
```{r}
plot(m_bty_gen$residuals ~ c(1:nrow(evals)))
```
    
    **Variability Across Groups:** Residuals against each predictor variable
    There are not any obvious deviations in variability across groups for either of the predictor variables (`bty_avg` & `gender`)
```{r}
plot(m_bty_gen$residuals ~ jitter(evals$bty_avg))
plot(m_bty_gen$residuals ~ evals$gender)
```
    
    From the plots above, the data appear to reasonably meet the conditions for this model.

8.  Is `bty_avg` still a significant predictor of `score`? Has the addition
    of `gender` to the model changed the parameter estimate for `bty_avg`?
    
    Yes, `bty_avg` is still a significant predictor of score and the addition of `gender` to the model improves the parameter estimates. However, the adjusted R^2 is still very low (5.5%), so our model still does not explain much of the data's variance.
    
    

Note that the estimate for `gender` is now called `gendermale`. You'll see this 
name change whenever you introduce a categorical variable. The reason is that R 
recodes `gender` from having the values of `female` and `male` to being an 
indicator variable called `gendermale` that takes a value of $0$ for females and
a value of $1$ for males. (Such variables are often referred to as "dummy" 
variables.)

As a result, for females, the parameter estimate is multiplied by zero, leaving 
the intercept and slope form familiar from simple regression.

\[
  \begin{aligned}
\widehat{score} &= \hat{\beta}_0 + \hat{\beta}_1 \times bty\_avg + \hat{\beta}_2 \times (0) \\
&= \hat{\beta}_0 + \hat{\beta}_1 \times bty\_avg\end{aligned}
\]

We can plot this line and the line corresponding to males with the following 
custom function.

```{r twoLines}
multiLines(m_bty_gen)
#summary(m_bty_gen)
```

9.  What is the equation of the line corresponding to males? (*Hint:* For
    males, the parameter estimate is multiplied by 1.) For two professors
    who received the same beauty rating, which gender tends to have the
    higher course evaluation score?
    
    Equation of the line corresponding to males:
    \[
  \begin{aligned}
\widehat{score} &= \hat{\beta}_0 + \hat{\beta}_1 \times bty\_avg + \hat{\beta}_2 \times (1) \\&= 3.74734 + 0.07416 \times bty\_avg + 0.17239 \times (1) \end{aligned}
\]

    Males tend to have a slightly higher course evaluation score.

The decision to call the indicator variable `gendermale` instead of`genderfemale`
has no deeper meaning. R simply codes the category that comes first 
alphabetically as a $0$. (You can change the reference level of a categorical 
variable, which is the level that is coded as a 0, using the`relevel` function. 
Use `?relevel` to learn more.)

10. Create a new model called `m_bty_rank` with `gender` removed and `rank` 
    added in. How does R appear to handle categorical variables that have more 
    than two levels? Note that the rank variable has three levels: `teaching`, 
    `tenure track`, `tenured`.
    
```{r}
m_bty_rank <- lm(score ~ bty_avg + rank, data = evals)
summary(m_bty_rank)
```
    
    From the summary of `m_bty_rank` we see that R handles categorical variables with more than 2 levels by giving multiple coefficients to describe the levels.

The interpretation of the coefficients in multiple regression is slightly 
different from that of simple regression. The estimate for `bty_avg` reflects
how much higher a group of professors is expected to score if they have a beauty
rating that is one point higher *while holding all other variables constant*. In
this case, that translates into considering only professors of the same rank 
with `bty_avg` scores that are one point apart.

## The search for the best model

We will start with a full model that predicts professor score based on rank, 
ethnicity, gender, language of the university where they got their degree, age, 
proportion of students that filled out evaluations, class size, course level, 
number of professors, number of credits, average beauty rating, outfit, and 
picture color.

11. Which variable would you expect to have the highest p-value in this model? 
    Why? *Hint:* Think about which variable would you expect to not have any 
    association with the professor score.
    
    I expect the variable `cls_profs` to have the highest p-value, because students rating a professor have an experience independant of other professors teaching a section of the same course and may even be unaware that other professors are teaching different sections.

Let's run the model...

```{r m_full}
m_full <- lm(score ~ rank + ethnicity + gender + language + age + cls_perc_eval 
             + cls_students + cls_level + cls_profs + cls_credits + bty_avg 
             + pic_outfit + pic_color, data = evals)
summary(m_full)
```

12. Check your suspicions from the previous exercise. Include the model output
    in your response.
    
```{r}
summ_full <- summary(m_full)
which.max(summ_full$coefficients[,4])
```
    
    `cls_profssingle` (`cls_profs`) appears to have the highest p-value (0.77806). My next guess for highest p-value was actually the valiable `pic_color` so I was suprised to see that it, in fact, evaluates to significant (p-val = 0.00252).

13. Interpret the coefficient associated with the ethnicity variable.

    When all other variables are held constant, non-minority professors recieved scores 0.123 higher than minority professors

14. Drop the variable with the highest p-value and re-fit the model. Did the
    coefficients and significance of the other explanatory variables change?
    (One of the things that makes multiple regression interesting is that
    coefficient estimates depend on the other variables that are included in
    the model.) If not, what does this say about whether or not the dropped
    variable was collinear with the other explanatory variables?
    
```{r}
m_no_cls_profs <- lm(score ~ rank + ethnicity + gender + language + age + cls_perc_eval 
             + cls_students + cls_level + cls_credits + bty_avg 
             + pic_outfit + pic_color, data = evals)
summary(m_no_cls_profs)
```
    
    After dropping `cls_profs` there was only a very slight change in the coefficients and significance of the model fits. This finding suggests that the variable `cls_profs` is collinear with the other variables and did not add much explanatory value (account for variance) to the model.

15. Using backward-selection and p-value as the selection criterion,
    determine the best model. You do not need to show all steps in your
    answer, just the output for the final model. Also, write out the linear
    model for predicting score based on the final model you settle on.
    
```{r, message=F}
library(olsrr)
#ls_step_best_subset(m_full)
m_backp <- ols_step_backward_p(m_full, details = T)
m_formula <- m_backp$model$terms
Reduce( paste, deparse(m_formula))


m_final <- lm(score ~ rank + ethnicity + gender + language + age + cls_perc_eval 
             + cls_students + cls_level + cls_credits + bty_avg 
             + pic_outfit + pic_color, data = evals)
```
    

16. Verify that the conditions for this model are reasonable using diagnostic 
    plots.
    
    **Model Diagnostics** (following 9.4.3 Checking model conditions using graphs)  
    **Nearly Normal Residuals:** Check for outliers on residuals histogram
    There are no obvious sign of 'major' outliers in the distribution
```{r}
hist( m_final$residuals )
```

    **Check for constant variance:** absolute value of residuals against the fitted vlue
    There are no obvious deviations from constant variance

```{r}
plot(abs(m_final$residuals) ~ jitter(m_final$fitted.values))
```

    **Data Independence:** Residuals in Order of their data collection
    There is no structure to the data as a function of data collection, so we can assume indepenence.
```{r}
plot(m_final$residuals ~ c(1:nrow(evals)))
```
    
    **Variability Across Groups:** Residuals against each predictor variable
    There are not any obvious deviations in variability across groups for either of the predictor variables (`bty_avg` & `gender`)
```{r}
plot(m_final$residuals ~ jitter(evals$bty_avg))
plot(m_final$residuals ~ evals$gender)
plot(m_final$residuals ~ evals$rank)
plot(m_final$residuals ~ evals$ethnicity)
plot(m_final$residuals ~ evals$language)
plot(m_final$residuals ~ jitter(evals$age))
plot(m_final$residuals ~ evals$cls_perc_eval)
plot(m_final$residuals ~ evals$cls_students)
plot(m_final$residuals ~ evals$cls_level)
plot(m_final$residuals ~ evals$cls_credits)
plot(m_final$residuals ~ evals$pic_outfit)
plot(m_final$residuals ~ evals$pic_color)
```
    
    From the plots above, the data appear to reasonably meet the conditions for this model. Although, for the variable cls_students there are several outliers, this is passable because the sample size is very large.
    
    
    
17. The original paper describes how these data were gathered by taking a
    sample of professors from the University of Texas at Austin and including 
    all courses that they have taught. Considering that each row represents a 
    course, could this new information have an impact on any of the conditions 
    of linear regression?
    
    Yes, this has an impact on independence of the data samples, because if the same professor has tought multiple classes then the data points generated for classes by the same professor are not indepent.

18. Based on your final model, describe the characteristics of a professor and 
    course at University of Texas at Austin that would be associated with a high
    evaluation score.
    
    To be associated with a high evaluation score, a UT@A professor would most likely be a young and beautiful non-minority male who is early on in his career (tenure track). He likely graduated from an English speaking university and teaches a one-credit upper level course. His picture was likely a formal gray-scale image.

19. Would you be comfortable generalizing your conclusions to apply to professors
    generally (at any university)? Why or why not?
    
    No, this was an observational study of classes at UT@A. There was no controlled experiment to investigate the relationship of average beauty rating and scores. Therefore, this cannot be generalized to all universities. However, it is an interesting observation that could be used as the basis of a controlled experiment to learn more about student bias for percieved beauty of thier professors.




<!-- This is a product of OpenIntro that is released under a Creative Commons Attribution-ShareAlike 3.0 Unported. This lab was written by Mine Çetinkaya-Rundel and Andrew Bray. -->
