---
title: 'Inference for numerical data'
output:
  pdf_document: default
  html_document:
    css: ./lab.css
    highlight: pygments
    theme: cerulean
---

## North Carolina births

In 2004, the state of North Carolina released a large data set containing 
information on births recorded in this state. This data set is useful to 
researchers studying the relation between habits and practices of expectant 
mothers and the birth of their children. We will work with a random sample of 
observations from this data set.

## Exploratory analysis

Load the `nc` data set into our workspace.

```{r load-data, eval=TRUE}
load("more/nc.RData")
dim( nc )
```

We have observations on 13 different variables, some categorical and some 
numerical. The meaning of each variable is as follows.

variable         | description
---------------- | -----------
`fage`           | father's age in years.
`mage`           | mother's age in years.
`mature`         | maturity status of mother.
`weeks`          | length of pregnancy in weeks.
`premie`         | whether the birth was classified as premature (premie) or full-term.
`visits`         | number of hospital visits during pregnancy.
`marital`        | whether mother is `married` or `not married` at birth.
`gained`         | weight gained by mother during pregnancy in pounds.
`weight`         | weight of the baby at birth in pounds.
`lowbirthweight` | whether baby was classified as low birthweight (`low`) or not (`not low`).
`gender`         | gender of the baby, `female` or `male`.
`habit`          | status of the mother as a `nonsmoker` or a `smoker`.
`whitemom`       | whether mom is `white` or `not white`.

1.  What are the cases in this data set? How many cases are there in our sample?

    Each case (row) of the data set is a record for a child's birth with the child's weight being 1 of 13 posible features for an observation. There are 1000 cases in this sample (from use of dim() function in previous code block).

As a first step in the analysis, we should consider summaries of the data. This
can be done using the `summary` command:

```{r summary, eval=TRUE}
summary(nc)
```

As you review the variable summaries, consider which variables are categorical 
and which are numerical. For numerical variables, are there outliers? If you 
aren't sure or want to take a closer look at the data, make a graph.

Consider the possible relationship between a mother's smoking habit and the 
weight of her baby. Plotting the data is a useful first step because it helps 
us quickly visualize trends, identify strong associations, and develop research
questions.

2.  Make a side-by-side boxplot of `habit` and `weight`. What does the plot 
highlight about the relationship between these two variables?

    In a side-by-side boxplot, we can see that median weight of babies born to non-smokers is slightly higher than the median weight of babies for to smokers. However, there is a lot of overlap between the two distributions.
    
```{r}
boxplot(nc$weight ~ nc$habit)
```
    
    

The box plots show how the medians of the two distributions compare, but we can
also compare the means of the distributions using the following function to 
split the `weight` variable into the `habit` groups, then take the mean of each
using the `mean` function.

```{r by-means, eval=TRUE}
by(nc$weight, nc$habit, mean)
```

There is an observed difference, but is this difference statistically 
significant? In order to answer this question we will conduct a hypothesis test
.

## Inference

3.  Check if the conditions necessary for inference are satisfied. Note that 
you will need to obtain sample sizes to check the conditions. You can compute 
the group size using the same `by` command above but replacing `mean` with 
`length`.

    We are working with a random sample, so the independence condition is met. From the output of the block below, we can see that both are greater than 30. Additionally, from the plot above we can see that there arent any particularly extreme outliers, therefore it is reasonable to assume the normality condition.
    
```{r, eval=TRUE}
by(nc$weight, nc$habit, length)
```
    

4.  Write the hypotheses for testing if the average weights of babies born to 
smoking and non-smoking mothers are different.

    **Null Hypothesis:** there is no difference between the mean weights of babaies born to smoking and non-smoking mothers  
    **Alternative Hypothesis:** there is a difference between the means for the weights of babies born to smoking and non-smoking mothers

Next, we introduce a new function, `inference`, that we will use for conducting
hypothesis tests and constructing confidence intervals. 

```{r inf-weight-habit-ht, eval=TRUE, tidy=FALSE}
inference(y = nc$weight, x = nc$habit, est = "mean", type = "ht", 
          null = 0, alternative = "twosided", method = "theoretical")
```

Let's pause for a moment to go through the arguments of this custom function. 
The first argument is `y`, which is the response variable that we are 
interested in: `nc$weight`. The second argument is the explanatory variable, 
`x`, which is the variable that splits the data into two groups, smokers and 
non-smokers: `nc$habit`. The third argument, `est`, is the parameter we're 
interested in: `"mean"` (other options are `"median"`, or `"proportion"`.) Next
we decide on the `type` of inference we want: a hypothesis test (`"ht"`) or a 
confidence interval (`"ci"`). When performing a hypothesis test, we also need 
to supply the `null` value, which in this case is `0`, since the null 
hypothesis sets the two population means equal to each other. The `alternative`
hypothesis can be `"less"`, `"greater"`, or `"twosided"`. Lastly, the `method` 
of inference can be `"theoretical"` or `"simulation"` based.

5.  Change the `type` argument to `"ci"` to construct and record a confidence 
interval for the difference between the weights of babies born to smoking and 
non-smoking mothers.

```{r}
inference(y = nc$weight, x = nc$habit, est = "mean", type = "ci", 
          null = 0, alternative = "twosided", method = "theoretical")
```


By default the function reports an interval for ($\mu_{nonsmoker} - \mu_{smoker}$)
. We can easily change this order by using the `order` argument:

```{r inf-weight-habit-ci, eval=TRUE, tidy=FALSE}
inference(y = nc$weight, x = nc$habit, est = "mean", type = "ci", 
          null = 0, alternative = "twosided", method = "theoretical",
          order = c("smoker","nonsmoker"))
```

* * *

## On your own

-   Calculate a 95% confidence interval for the average length of pregnancies 
(`weeks`) and interpret it in context. Note that since you're doing inference 
on a single population parameter, there is no explanatory variable, so you can 
omit the `x` variable from the function.

    From the console output below, we find the 95 % Confidence interval = ( 38.1528 , 38.5165 ).

```{r}
inference(y = nc$weeks, est = "mean", type = "ci", 
          null = 0, alternative = "twosided", method = "theoretical")
```


-   Calculate a new confidence interval for the same parameter at the 90% 
confidence level. You can change the confidence level by adding a new argument 
to the function: `conflevel = 0.90`.

    From the console output below, we find the 90% Confidence interval = ( 38.182 , 38.4873 )
```{r}
inference(y = nc$weeks, est = "mean", type = "ci", conflevel = 0.9, 
          null = 0, alternative = "twosided", method = "theoretical")
```


-   Conduct a hypothesis test evaluating whether the average weight gained by 
younger mothers is different than the average weight gained by mature mothers.

    * State our hypotheses:  
    **Null Hypothesis:** there is no difference between the mean weights of babaies born to smoking and non-smoking mothers  
    **Alternative Hypothesis:** there is a difference between the means for the weights of babies born to smoking and non-smoking mothers  

    * See if the conditions of inference are met: we already established that the sample is random. From the output below, we see that the number of samples for each conditions is >30 and from the boxplot we do not observe any paticularly extreme outliers. Therefore, the normality condition is met so we can proceed with testing our hypothesese
```{r, eval=TRUE}
by(nc$gained, nc$mature, length)
```
```{r, eval=TRUE}
boxplot(nc$gained ~ nc$mature)
```
    
    * we will use an alpha = 0.05
```{r}
inference(y = nc$gained, x = nc$mature, est = "mean", type = "ht",
          conflevel = 0.95, null = 0, alternative = "twosided", 
          method = "theoretical")
```
    
    From the output above, we see that our hypothesis test yields a p-value of 0.1686.  
    This value is much larger than our alpha value.  
    Therefore we accept the null hypothesis which states that there is no difference of  
    weight gain between young and mature mothers.

-   Now, a non-inference task: Determine the age cutoff for younger and mature 
mothers. Use a method of your choice, and explain how your method works.

    We can retrieve the stats derived for the bosplot below & we see that the minimum value for the 'mature mom' is 35 while the max value for 'younger mom' is 34. Therefore, we find:  
    
    * **mature mom** >= 35 years
    * **younger mom** <= 34 years

```{r}
bp <- boxplot(nc$mage ~ nc$mature)
bp$stats
```


-   Pick a pair of numerical and categorical variables and come up with a 
research question evaluating the relationship between these variables. 
Formulate the question in a way that it can be answered using a hypothesis test
and/or a confidence interval. Answer your question using the `inference` 
function, report the statistical results, and also provide an explanation in 
plain language.

    **Question:** Is the average age of mothers to premature children different from the average age of mothers to full-term children?  
    
    State our hypotheses:  
    **Null Hypothesis:** There is no difference between the mean ages of mothers to premie vs full-term babies.  
    **Alternative Hypothesis:** There exists a difference between the mean ages of mothers to premie vs full-term babies.  
    
    Conditions for Inference:  
    **Independence** Yes, this is a random sample  
    **Normality** Yes, from the output below we can see that both conditions have n>30 and the boxplot shows that there are no particularly extreme outliers
    
```{r, eval=TRUE}
by(nc$mage, nc$premie, length)
```
```{r, eval=TRUE}
boxplot(nc$mage ~ nc$premie)
```

    The conditions of inference are met, so we can move forward and will use an alpha = 0.05
    
```{r}
inference(y = nc$mage, x = nc$premie, est = "mean", type = "ht",
          conflevel = 0.95, null = 0, alternative = "twosided", 
          method = "theoretical")
```
    
    * Our test resulted in a very large p-value.  
    Therefore we accept the null hypothesis which states that there is no difference between the mean  
    ages of mothers who gave birth to premies and mothers who gave birth to full-term babies.


<!-- This is a product of OpenIntro that is released under a Creative Commons Attribution-ShareAlike 3.0 Unported. This lab was written by Mine Çetinkaya-Rundel and Andrew Bray. -->
